\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-03-12 papers/methods/paper.tex
Updating caption
Initial draft feedback

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account full analysis systematics, Malmquist bias, underlying populations and standardisation coefficients. Our model is optimised for efficient fitting and has been validated on a combined dataset of more than $250\,000$ supernovae. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations quantify the bias for Flat $w$CDM cosmology, with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ included to allow rigorous determination of model bias in $w$. Bias in SNANA tests across intrinsic dispersion models is found to range from $-0.15\sigma$ to $+0.27\sigma$ in $w$ for datasets of comparable strength to the Dark Energy Survey three-year analysis. 
\end{abstract}

@@

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, of the number of observed Type Ia supernovae (SN Ia) have increased by more than an order of magnitude thanks to modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Balland2009, Amanullah2010}. Cosmological analysis of these supernova samples \citep{Kowalski2008, Conley2011, Suzuki2012, Betoule2014, Rest2014} have been combined with complimentary probes of large scale structure \citep{Alam2017} and the CMB \citep{Hinshaw2013, PlanckCollaboration2013}. Despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

@@

As such, staggering effort is being put into developing more sophisticated supernovae analyses. First used in ESSENSE analyses \citep{Wood-Vasey2007}, and then refined and improved in \citet{Kessler2009}, the role of simulations has become increasingly important in determining biases and validating models. \citet{Betoule2014} make use of redshift dependent $\mu$-bias to correct their analyses, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulation corrections as a function of redshift, stretch and colour to better remove fitting biases. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with the cost of increased computational time \citep{Weyant2013, Jennings2016}. Hierarchical Bayesian models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, however often face difficulties finding sufficient analytic approximations for complicated effects such as Malmquist bias.

@@

The underlying actual supernova population is not as clear cut, and indeed accurately characterising this population, its evolution over redshift and effects from environment is one of the challenges of supernova cosmology. However, given some modelled underlying population that lives in the redshift-dependent space $M_B$, $x_1$ and $c$, the introduction of cosmology into the model is simple -- it is encoded in the functional map between those two populations, from apparent magnitude space to absolute magnitude. Specifically, for any given supernova our functional map may take the traditional form:
\begin{equation}
M_B = m_B + \alpha x_1 - \beta c - \mu(z) + \text{corrections}, \label{eq:standard}
\end{equation}
where $\alpha$ is the stretch correction \citep{Phillips1993}, and $\beta$ is the colour correction \citep{Tripp1998} that respectively encapsulate the empirical relation that broader and bluer supernovae are brighter. The $\text{corrections}$ term at the end often includes corrections for host galaxy environment, as this has statistically significant effects on supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013, Uddin2017}. The cosmological term, $\mu(z)$ represents the distance modulus, and is precisely known given cosmological parameters and an input redshift.

@@

Traditional $\chi^2$ analyses such as that found in \citet{Kowalski2008, Conley2011, Betoule2014}, minimise the difference in distance modulus between the observed distance modulus $\mu_{\rm obs}$ (attained from rearranging equation \eqref{eq:standard}), and the cosmologically predicted values $\mu_C$; given by
\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right]\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}} \label{eq:mu0}\\
\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B. \label{eq:mu1}
\end{align}
The minimising function is then given as
\begin{equation}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1} (\mu_{\rm obs} - \mu_C)
\end{equation}
where $C^{-1}$ is an uncertainty matrix that combines the uncertainty from the SALT2 fits, intrinsic dispersion, calibration, dust, peculiar velocity and many other factors (see \citet{Betoule2014} for a review). The benefit this analysis methodology provides is speed - for samples of hundreds of supernovae, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with two costs. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of uncertainty Gaussianity. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus loses all information about uncertainty correlation. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

@@

Sitting comfortably between the traditional models simplicity and the complexity of forward modelling lies hierarchical Bayesian Models. With the introduction of multiple layers in our model, we can add far more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009}, \citet{Mandel2011} and \citet{Mandel2017} construct a hierarchical model that they apply to the light curve fitting for supernova. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide a model that offers increased flexibility with reduced uncertainty over the traditional method, but does not incorporate bias correction. \citet{March2014, Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS like simulations. The recent BAHAMAS model \citep{Shariff2016} builds off this and reanalyses the JLA dataset (using their redshift dependent bias corrections), whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} also performed a reanalysis of the JLA dataset, finding significant differences in $\alpha$ and $\beta$ values from the original as well by correctly including in their likelihood the determinant of the global covariance matrix. Notably, these methods rely on data that is bias corrected or ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The simplification made by the UNITY analysis is that the bias is well described by an analytic function, and this function is fixed such that selection effects are assumed to be determined with perfect precision. Their model was validated on thirty realisations of statistics. Whilst there are a large amount of hierarchical models available, all sofar mentioned, in the case where any validation has been done, have been done on either $\Lambda$CDM cosmology or Flat $\Lambda$CDM cosmology. None of them have undergone high-statistics simulation verification for $w$CDM cosmology to quantify each models respective bias, which is becoming critically important as precision supernovae cosmology comes into its own and focus shifts from determination of $\Omega_m$ to $w$.

@@

We construct our hierarchical Bayesian model with several goals in mind: creation of a redshift-dependent underlying supernova population, increased treatment of systematics, and analytic correction of selection effects, including systematic uncertainty. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in Stan \citep{Carpenter2017, StanDevelopmentTeam2017} which uses automatic differentiation and the no-U-turn Sampler (NUTS), which is a variant of Hamiltonian Monte Carlo, to efficiently sample high dimensional parameter space.

@@

Like most of the BHM methods introduced previously, we work from the summary statistics as well, where each observed supernova has an apparent magnitude $\hat{m_B}$, stretch $\hat{x_1}$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. Given machine learning classifiers to type supernovae, we will also have a probability of being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m_B}, \hat{x_1}, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

The first layer of the hierarchy represents the latent population of $m_B$, $x_1$ and $c$. This can be thought of as the `true' values for the supernova, and with correct summary statistics and uncertainty would be normally distributed around the observed values such that 
\begin{equation}
P(\hat{m_B},\hat{x_1},\hat{c}|m_B, x_1, c,\cov) = \mathcal{N}(\lbrace \hat{m_B},\hat{x_1},\hat{c} \rbrace|\lbrace m_B, x_1, c \rbrace, \cov). \label{eq:pop}
\end{equation}
However, the presence of chromatic smearing in the intrinsic supernova population \citep{Guy2010, Chotard2011} means the summary statistics report from light curve fitting are not Gaussian. We model the extra dispersion included in the observations from colour smearing by adding extra independent uncertainty on the colour observation. As shown in \citep{Kessler2013}, the extra dispersion shows heavy redshift dependence. As this may represent underlying physics more than the effects of Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $\kappa_0 + \kappa_1 z$ to our observed colour uncertainty (in quadrature), where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to enforce minimum possible extra uncertainty in our model. Each survey has its own set of $\kappa$ parameters. Furthermore, we use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the bias in the mean colour measured as a function of redshift, and subtract this bias from our observed data. We allow the amount of bias to be parametrised to account for different intrinsic scatter models, where parameter $s_c$ controlled the fraction of bias subtraction - 0 for no bias, up to 1 for the maximum bias found in simulations (which corresponded to the intrinsic scatter model from \citet{Chotard2011}). Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a universal shift in supernovae biases. We place a uniform prior on $s_c$.

As we are focused on the spectroscopically confirmed supernovae for this iteration of the method, we assume the observed redshift $\hat{z}$ is the true redshift $z$ such that $P(\hat{z}|z) = \delta(\hat{z} - z)$. Similarly, we take the mass probability estimate $\hat{m}$ as correct, and do not model a latent variable.

\begin{figure}
\begin{tikzpicture}
% Obs nodes
\node[obs]  (obssum) {$\hat{m_B}, \hat{x_1}, \hat{c}, \mathcal{C}^{-1}$};
\node[obs, left=0.8cm of obssum]  (zobs) {$\hat{z}$};
\node[obs, right=0.8cm of obssum]  (mobs) {$\hat{m}$};
\node[obs, right=0.8cm of mobs]  (pobs) {$\hat{p}$};

@@

% Plates
\plated[thick] {sn} {(obssum)(zobs)(mobs)(pobs)(sum)(z)(m)(mu)} {$\forall\ \rm{SN}$} ;
\plated[thick] {survey} {(sn)(pop1)} {\\$\forall\ \rm{Survey}$} ;
\end{tikzpicture}
\caption{Probabilistic graphical model for our likelihood. Double-lined nodes represent observed variables, diamond nodes represent deterministic variables, and standard nodes represent fit variables.}
\label{fig:pgm}
\end{figure}

@@

The underlying supernova population is characterised by underlying distributions in colour, stretch and magnitude. Analytic models often treat the colour and stretch populations with a skew normal and normal, respectively, and have them as independent. Intrinsic dispersion is treated in a variety of manners in other models, from representing it simply as (Gaussian) scatter on the absolute magnitude of the supernova population, to correlated multivariate normal scatter on the combined magnitude, stretch and colour distribution. Additionally, redshift drift of populations' mean colour and stretch will introduce a cosmological bias in our fits unless the population possesses similar ability to change as a function of redshift.

@@

\subsubsection{Uncertainty Propagation}
\label{sec:systreat}
The chief difficulty with including systematic uncertainties in supernova analyses is that they generally occur during the observational pipeline, and have difficult-to-model effects on the output observations. As such, the normal treatment for systematics is to compute their effect on the supernova summary statistics -- computing the numerical derivatives $\frac{\partial \hat{m_B}}{\partial \Z_i}$, $\frac{\partial \hat{x_1}}{\partial \Z_i}$, $\frac{\partial \hat{c}}{\partial \Z_i}$, where $\Z_i$ represents the $i$\textsuperscript{th} systematic.

@@

where the denominator represents an integral over all potential data. Full derivation of this can be found in Appendix \ref{app:selection}. As $\theta$ represents the vector of all model parameters, and $D$ represents a vector of all observed variables, this is not a trivial integral. Techniques to approximate this integral, such as Monte-Carlo integration or high dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by HMC, and post-fitting importance sampling failed due to high-dimensionality (a brief dismissal of many months of struggle). We therefore simplify the integral and approximate the selection effects in apparent magnitude and redshift space independently, such that the denominator of equation \eqref{eq:main}, denoted now $w$ for simplicity, is given as

@@

We apply two further approximations similar to those made in {\rubin} -- that the redshift distribution of the observed supernova reasonably well sampled the $P(S|z)P(z|\theta)$ distribution, and that the survey colour and stretch populations can be treated as Gaussian for the purposes of this integral. It was found that discarding the skewness entirely resulted in highly biased population recovery, and so we instead characterise the skew normal colour distribution with a Gaussian that follows the mean and variance of a skew normal; with mean given by $\langle c(z) \rangle + \sqrt{\frac{2}{\pi}} \sigma_c \delta_c$ and variance $\sigma_c^2(1 - 2\delta_c^2/\pi)$. This shifted Gaussian approximation removes the unintended bias when simply discarding skewness. More detail on this shift can be found in Appendix \ref{app:approx}. The population $P(m_B | z, \theta)$ becomes $\mathcal{N}(m_B|m_B^*(z), \sigma^*_{m_B})$, where

@@

\sigma^*_{m_B} &= \sigma_{M_B}^2 + (\alpha \sigma_{x_1})^2 + (\beta \sigma_c)^2

@@

P(S|m_B) = \Phi^C(m_B | \mu_{\rm CDF}, \sigma_{\rm CDF}),

@@

where $\Phi^c$ the complimentary CDF and $\mu_{\rm CDF}$ and $\sigma_{\rm CDF}$ specify the selection function. The appropriateness of an error function has been found by many past surveys \citep{Dilday2008, Barbary2010, Perrett2012, Graur2013, Rodney2014}. However, for surveys which suffer from saturation and thus rejection of low-z supernovae, or for groups of surveys treated together (as is common to do with low-redshift surveys), we find that a skew normal is a suitable analytic form, taking the form

@@

The selection functions are fit to apparent magnitude efficiency ratios calculated from SNANA simulations, by calculating an efficiency ratio as a function of apparent magnitude. Uncertainty of the Malmquist bias from the imperfection of the analytic function is incorporated into the fit. Uncertainty was uniformly added to the efficiency ratio data from our simulations until the reduced $\chi^2$ of the analytic fit reached $1$, allowing us to extract an uncertainty covariance matrix for our analytic fits to either the error function or the skew normal.

With the well sampled approximation as specified previously, we can remove the redshift integral in Eq \eqref{eq:w1} and replace it with a correction for each observed supernova. For the error function (denoted with the subscript `CDF') and skew normal selection functions respectively (denoted with a subscript `Skew'), this correction becomes
\begin{align}
w_{\rm CDF} &= \Phi^c\left(  \frac{m^*_B - \mu_{\rm CDF}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm CDF}^2}}  \right) \label{eq:seldes}\\
w_{\rm Skew} &= 2\mathcal{N}\left( \frac{m_B^* - \mu_{\rm Skew}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm Skew}^2}}\right) \times \notag\\
&\quad\quad \Phi\left(  \frac{{\rm sign}(\alpha_{\rm Skew})(m_B^* - \mu_{\rm Skew})}{\frac{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}{\sigma^2_{\rm Skew}} \sqrt{\frac{\sigma^2_{\rm Skew}}{\alpha^2_{\rm Skew}} + \frac{\sigma_{m_B}^{*2} \sigma^2_{\rm Skew}}{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}} }\right), \label{eq:sellowz}
\end{align}
and is incorporated into our likelihood.

>>>>

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account a full range of analysis systematics and selection effects. This advances the work of \citet{Rubin2015} by including additional sources of systematic uncertainty and improved treatment of Malmquist bias whilst increasing numerical efficiency. Our model is optimised for efficient fitting and has been validated on a combined simulated dataset of more than $250\,000$ supernovae. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations quantify the bias for Flat $w$CDM cosmology, with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ included to allow rigorous determination of model bias in $w$. SNANA simulations can recover the input cosmology to within $0.27\sigma$, while taking into account many more effects than standard analyses.
\end{abstract}
% Bias in SNANA tests across intrinsic dispersion models is found to range from $-0.15\sigma$ to $+0.27\sigma$ in $w$ for datasets of comparable strength to the Dark Energy Survey three-year analysis.

@@

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed Type Ia supernovae (SN Ia) have increased by more than an order of magnitude thanks to modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Balland2009, Amanullah2010}. Cosmological analysis of these supernova samples \citep{Kowalski2008, Conley2011, Suzuki2012, Betoule2014, Rest2014} have been combined with complimentary probes of large scale structureand the CMB. For a recent review, see \citet{Huterer2018}. Despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

@@

As such, staggering effort is being put into developing more sophisticated supernovae analyses. First used in ESSENCE analyses \citep{Wood-Vasey2007}, and then refined and improved in \citet{Kessler2009}, the role of simulations has become increasingly important in determining biases and validating models. \citet{Betoule2014} make use of redshift dependent $\mu$-bias to correct their analyses, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulation corrections as a function of redshift, stretch and colour to better remove fitting biases. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with the cost of increased computational time \citep{Weyant2013, Jennings2016}. Hierarchical Bayesian models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, however often face difficulties finding sufficient analytic approximations for complicated effects such as Malmquist bias.

@@

The underlying actual supernova population is not as clear cut, and indeed accurately characterising this population, its evolution over redshift and effects from environment is one of the challenges of supernova cosmology. However, given some modelled underlying population that lives in the redshift-dependent space $M_B$ (absolute magnitude), $x_1$ and $c$, the introduction of cosmology into the model is simple -- it is encoded in the functional map between those two populations, from apparent magnitude space to absolute magnitude. Specifically, for any given supernova our functional map may take the traditional form:
\begin{equation}
M_B = m_B + \alpha x_1 - \beta c - \mu(z) + \text{corrections}, \label{eq:standard}
\end{equation}
where $\alpha$ is the stretch correction \citep{Phillips1993, Phillips1999}, and $\beta$ is the colour correction \citep{Tripp1998} that respectively encapsulate the empirical relation that broader and bluer supernovae are brighter. The $\text{corrections}$ term at the end often includes corrections for host galaxy environment, as this has statistically significant effects on supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013, Uddin2017}. The cosmological term, $\mu(z)$ represents the distance modulus, and is precisely known given cosmological parameters and an input redshift.

@@

Traditional $\chi^2$ analyses such as that found in \citet{Riess1998, Perlmutter1999, Wood-Vasey2007, Kowalski2008, Kessler2009, Conley2011, Betoule2014}, minimise the difference in distance modulus between the observed distance modulus $\mu_{\rm obs}$ (attained from rearranging equation \eqref{eq:standard}), and the cosmologically predicted values $\mu_C$. This function is given by
\begin{equation}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1} (\mu_{\rm obs} - \mu_C),
\end{equation}
where $C^{-1}$ is an uncertainty matrix that combines the uncertainty from the SALT2 fits, intrinsic dispersion, calibration, dust, peculiar velocity and many other factors (see \citet{Betoule2014} for a review). The cosmological $\mu_C$ is defined as
\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right]\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}} \label{eq:mu0}\\
\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B. \label{eq:mu1}
\end{align}

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with two costs. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of uncertainty Gaussianity. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about uncertainty correlation. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

@@

Sitting comfortably between the traditional models simplicity and the complexity of forward modelling lies hierarchical Bayesian Models. With the introduction of multiple layers in our model, we can add far more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009}, \citet{Mandel2011} and \citet{Mandel2017} construct a hierarchical model that they apply to the light curve fitting for supernova. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide a model that offers increased flexibility with reduced uncertainty over the traditional method, but does not incorporate bias correction. \citet{March2014, Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS like simulations. The recent BAHAMAS model \citep{Shariff2016} builds off this and reanalyses the JLA dataset (using their redshift dependent bias corrections), whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} also performed a reanalysis of the JLA dataset, finding significant differences in $\alpha$ and $\beta$ values from the original as well by correctly including in their likelihood the determinant of the global covariance matrix. Notably, these methods rely on data that is bias corrected or ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The simplification made by the UNITY analysis is that the bias is well described by an analytic function, and this function is fixed such that selection effects are assumed to be determined with perfect precision. Their model was validated on thirty realisations of statistics.

Whilst there are a large amount of hierarchical models available, all so far mentioned, in the case where any validation has been done, have been done on either $\Lambda$CDM cosmology or Flat $\Lambda$CDM cosmology. None of them have undergone high-statistics simulation verification for $w$CDM cosmology to quantify each models respective bias, which is becoming critically important as precision supernovae cosmology comes into its own and focus shifts from determination of $\Omega_m$ to $w$.

@@

We construct our hierarchical Bayesian model with several goals in mind: creation of a redshift-dependent underlying supernova population, increased treatment of systematics, and analytic correction of selection effects, including systematic uncertainty on those corrections. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in Stan \citep{Carpenter2017, StanDevelopmentTeam2017} which uses automatic differentiation and the no-U-turn Sampler (NUTS), which is a variant of Hamiltonian Monte Carlo, to efficiently sample high dimensional parameter space.

@@

Like most of the BHM methods introduced previously, we work from the summary statistics, where each observed supernova has an apparent magnitude $\hat{m}_B$, stretch $\hat{x}_1$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. We will also have a probability of being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

The first layer of the hierarchy represents the latent population of $m_B$, $x_1$ and $c$. This can be thought of as the `true' values for the supernova, and with correct summary statistics and uncertainty would be normally distributed around the observed values such that 
\begin{equation}
P(\hat{m}_B,\hat{x}_1,\hat{c}|m_B, x_1, c,\cov) = \mathcal{N}(\lbrace \hat{m}_B,\hat{x}_1,\hat{c} \rbrace|\lbrace m_B, x_1, c \rbrace, \cov). \label{eq:pop}
\end{equation}
However, the presence of chromatic smearing in the intrinsic supernova population \citep{Guy2010, Chotard2011} means the summary statistics reported from light curve fitting are not Gaussian. We model the extra dispersion included in the observations from colour smearing by adding extra independent uncertainty on the colour observation. As shown in \citep{Kessler2013}, the extra dispersion shows heavy redshift dependence. As the extra dispersion may arise from underlying supernova physics, rather than Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $\kappa_0 + \kappa_1 z$ to our observed colour uncertainty (in quadrature), where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to enforce minimum possible extra uncertainty in our model. The value of this prior was determined from running many iterations with differing prior strengths on our simple data set with simulated extra dispersion according to the results of \citet{Kessler2013}. Each survey has its own set of $\kappa$ parameters. Furthermore, we use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the shift in the mean colour measured as a function of redshift, and subtract this shift from our observed data. We allow the amount of shift to be parametrised to account for different intrinsic scatter models, where parameter $s_c$ controlled the fraction of bias subtraction -- 0 for no shift, up to 1 for the maximum shift found in simulations \citep[which corresponded to the intrinsic scatter model from][]{Chotard2011}. Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a universal shift in supernovae colour. We place a uniform prior on $s_c$.

As we are focused on the spectroscopically confirmed supernovae for this iteration of the method, we assume the observed redshift $\hat{z}$ is the true redshift $z$ such that $P(\hat{z}|z) = \delta(\hat{z} - z)$. Similarly, we take the mass probability estimate $\hat{m}$ as correct, and do not model a latent variable. One of the strengths of this model (and the {\rubin} analysis) is that for future data sets where supernovae have been classified photometrically, and we expect some misclassification and misidentification of the host galaxies, those can naturally be modelled and taken into account.

\begin{figure}
\begin{tikzpicture}
% Obs nodes
\node[obs]  (obssum) {$\hat{m}_B, \hat{x}_1, \hat{c}, \mathcal{C}^{-1}$};
\node[obs, left=0.8cm of obssum]  (zobs) {$\hat{z}$};
\node[obs, right=0.8cm of obssum]  (mobs) {$\hat{m}$};
\node[obs, right=0.8cm of mobs]  (pobs) {$\hat{p}$};

@@

% Plates
\plated[thick] {sn} {(obssum)(zobs)(mobs)(pobs)(sum)(z)(m)(mu)} {$\forall\ \rm{SN}$} ;
\plated[thick] {survey} {(sn)(pop1)} {\\$\forall\ \rm{Survey}$} ;
\end{tikzpicture}
\caption{Probabilistic graphical model for our likelihood. Double-lined nodes represent observed variables, diamond nodes represent deterministic variables, and standard nodes represent fit variables. The SN box represents observed variables and latent variables for each individual supernova, whilst the survey box represents survey-specific variables, which in general describe the supernovae population for the survey and the systematics associated with it.}
\label{fig:pgm}
\end{figure}

@@

The underlying supernova population is characterised by underlying distributions in colour, stretch and magnitude. Analytic models often treat the colour and stretch populations with a skew normal and normal, respectively, and have them as independent. Intrinsic dispersion is treated in a variety of manners in other models, from representing it simply as (Gaussian) scatter on the absolute magnitude of the supernova population, to correlated multivariate normal scatter on the combined magnitude, stretch and colour distribution. Additionally, redshift drift of populations' mean colour and stretch will introduce a cosmological bias in our fits unless the model population possesses similar ability to change as a function of redshift.

@@

\subsubsection{Uncertainty Propagation}
\label{sec:systreat}
The chief difficulty with including systematic uncertainties in supernova analyses is that they generally occur during the observational pipeline, and have difficult-to-model effects on the output observations. As such, the normal treatment for systematics is to compute their effect on the supernova summary statistics -- computing the numerical derivatives $\frac{\partial \hat{m}_B}{\partial \Z_i}$, $\frac{\partial \hat{x}_1}{\partial \Z_i}$, $\frac{\partial \hat{c}}{\partial \Z_i}$, where $\Z_i$ represents the $i$\textsuperscript{th} systematic.

@@

where the denominator represents an integral over all potential data. Full derivation of this can be found in Appendix \ref{app:selection}. As $\theta$ represents the vector of all model parameters, and $D$ represents a vector of all observed variables, this is not a trivial integral. Techniques to approximate this integral, such as Monte-Carlo integration or high dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by HMC, and post-fitting importance sampling failed due to high-dimensionality (a brief dismissal of many months of struggle). We therefore simplify the integral and approximate the selection effects from their full expression in all of $\theta$-space, to apparent magnitude and redshift space independently, such that the denominator of equation \eqref{eq:main}, denoted now $w$ for simplicity, is given as

@@

We now apply two further approximations similar to those made in {\rubin} -- that the redshift distribution of the observed supernova reasonably well sampled the $P(S|z)P(z|\theta)$ distribution, and that the survey colour and stretch populations can be treated as Gaussian for the purposes of this integral. It was found that discarding the skewness entirely resulted in highly biased population recovery, and so we instead characterise the skew normal colour distribution with a Gaussian that follows the mean and variance of a skew normal; with mean given by $\langle c(z) \rangle + \sqrt{\frac{2}{\pi}} \sigma_c \delta_c$ and variance $\sigma_c^2(1 - 2\delta_c^2/\pi)$. This shifted Gaussian approximation removes the unintended bias when simply discarding skewness. More detail on this shift can be found in Appendix \ref{app:approx}. The population $P(m_B | z, \theta)$ becomes $\mathcal{N}(m_B|m_B^*(z), \sigma^*_{m_B})$, where

@@

\sigma^*_{m_B} &= \sigma_{M_B}^2 + (\alpha \sigma_{x_1})^2 + (\beta \sigma_c)^2.

@@

P(S|m_B) = \Phi^c(m_B | \mu_{\rm CDF}, \sigma_{\rm CDF}),

@@

where $\Phi^c$ the complimentary CDF and $\mu_{\rm CDF}$ and $\sigma_{\rm CDF}$ specify the selection function. The appropriateness of an error function has been found by many past surveys \citep{Dilday2008, Barbary2010, Perrett2012, Graur2013, Rodney2014}. However, for surveys which suffer from saturation and thus rejection of low-redshift supernovae, or for groups of surveys treated together (as is common to do with low-redshift surveys), we find that a skew normal is a suitable analytic form, taking the form

@@

The selection functions are fit to apparent magnitude efficiency ratios calculated from SNANA simulations, by calculating an efficiency ratio as a function of apparent magnitude. Uncertainty of the Malmquist bias from the imperfection of the analytic function is incorporated into the fit. Uncertainty was uniformly added in quadrature (similar to adding intrinsic dispersion as a fit parameter) to the efficiency ratio data from our simulations until the reduced $\chi^2$ of the analytic fit reached $1$, allowing us to extract an uncertainty covariance matrix for our analytic fits to either the error function or the skew normal.

With the well sampled approximation we can remove the redshift integral in Eq \eqref{eq:w1} and replace it with a correction for each observed supernova. For the error function (denoted with the subscript `CDF') and skew normal selection functions respectively (denoted with a subscript `Skew'), this correction becomes
\begin{align}
w_{\rm CDF} &= \Phi^c\left(  \frac{m^*_B - \mu_{\rm CDF}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm CDF}^2}}  \right) \label{eq:seldes}\\
w_{\rm Skew} &= 2\mathcal{N}\left( \frac{m_B^* - \mu_{\rm Skew}}{\sqrt{\sigma^{*2}_{m_B} + \sigma_{\rm Skew}^2}}\right) \times \notag\\
&\quad\quad \Phi\left(  \frac{{\rm sign}(\alpha_{\rm Skew})(m_B^* - \mu_{\rm Skew})}{\frac{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}{\sigma^2_{\rm Skew}} \sqrt{\frac{\sigma^2_{\rm Skew}}{\alpha^2_{\rm Skew}} + \frac{\sigma_{m_B}^{*2} \sigma^2_{\rm Skew}}{\sigma_{m_B}^{*2} + \sigma^2_{\rm Skew}}} }\right), \label{eq:sellowz}
\end{align}
and is incorporated into our likelihood.
